Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-04-27 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
add conclusion and study area text

The impacts of food insecurity during the pandemic are alarming, since diet-related diseases, such as obesity, heart-disease, and diabetes, were already critical public health concerns in Canada prior to COVID-19 [@boucher2017ontario]. While food banks are not necessarily a stable solution to food insecurity and in fact may encourage a retrenchment of neoliberal policy [@wakefield2013sweet], at least they can be argued to provide a resource of last instance to households in precarious situations [@bazerghi2016role]. As recently as 2019, the Hamilton Hunger Report^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf] noted that food banks in Hamilton, Ontario, recorded the highest number of visitors in the past 29 years; the number of children visiting food banks (minors up to 18 years old) was 9,125 in March 2019, up from 8,278 the year before, a rate of increase greater than population growth.

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days [@statisticscanada2020food]. Considering the negative mental and physical health effects associated with food insecurity, increases in food insecurity rates due to the pandemic, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early]. Recent data^[https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis_Exec-Sum_EN.pdf] from FBC showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary, which suggests that many households may again turn to food bank services to meet their needs.

@@

**Need to provide some context here: the geography which makes transportation challenging, i.e., due to the escarpment; the main parts of the city: city core, inner suburbs, industrial north, suburban, exurban; and the industrial background of the city maybe, with poverty rates compared to other Canadian cities?**

@@

The locations of food banks and related food services were obtained from the Hamilton Public Library's Food Access Guide^[http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf]. The guide was updated in April of 2021 to indicate any change affected on the services due to the pandemic. This includes modified business hours, a need to make reservations before frequenting, and locations that have completely shut down in consequence. Table 1 defines each service type and the number of locations pre- and during the COVID-19 pandemic. While some food bank services have a specific a target population, such as prioritizing families with young children aged between 0 and 3 or accepting only those providing proof of low-income status through housing and utility statements, all the food bank services indicated below are designed to accommodate those in need of food at zero to low cost. With our focus on food banks and food services that offer free or low-cost meals at particular locations, we first removed services such as Meals on Wheels and other food access services such as food box, community kitchens, student nutrition programs, and shopping and transportation. In addition, two free meal services held on different days at the same location were collapsed into a single service point for the accessibility analysis. Additional details on the operations of individual facilities is not publicly available, and with the changes in operations it proved unfeasible to collect it. For this reason, the analysis to follow is of accessibility to the location of food banks and services, but not to specific services (e.g., breakfasts vs. food boxes).

@@

Words go here.

We use data from the census to model low-income households as a proxy for households that are possibly experiencing food insecurity. However, the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known.

While the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown as approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations.

In the absence of information on how food insecure households travel to food banks and related services, we adopted a normative approach that examines accessibility to food banks reachable within 15 minutes based on standards for accessible drinking water...

>>>>

The impacts of food insecurity during the pandemic are alarming, since diet-related diseases, such as obesity, heart-disease, and diabetes, were already critical public health concerns in Canada prior to COVID-19 [@boucher2017ontario]. While food banks are not necessarily a stable solution to food insecurity and in fact may encourage a retrenchment of neoliberal policy [@wakefield2013sweet], at least they can be argued to provide a resource of last instance to households in precarious situations [@bazerghi2016role]. As a mid-size city grappling with deindustrialization, Hamilton exhibits high rates of poverty and food bank use. As recently as 2019, the Hamilton Hunger Report^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf] noted that food banks in the city recorded the highest number of visitors in the past 29 years; a rate of increase greater than population growth. Most troubling, approximately 40% of all visitors were children.

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days [@statisticscanada2020food]. Considering the negative mental and physical health effects associated with food insecurity, increases in food insecurity rates due to the pandemic, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early]. Recent data^[https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis_Exec-Sum_EN.pdf] from FBC showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary. Although more recent statistics on food bank use in Hamilton in 2020 and 2021 is not yet available, data from the Daily Bread Food Bank in neighbouring Toronto for August 2020 shows visits climbing 51% year-over-year in that city^[https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf], which suggests that many households in Hamilton are likely to turn to food bank services to meet their needs.

@@

With a population of around 540,000, the City of Hamilton is the fourth largest city in Ontario. It has historically been home to major manufacturing industries but de-industrialization that has occurred over the past several decades has led Hamilton to become one of the most highly divided cities in Ontario, with a significant proportion of its residents sustaining livings at Canada’s poverty level [@deluca2012code; @jakar2019turning; @latham2007determinants]. In 2016, the Hamilton Community Foundation reported that based on the Low-Income Cut-Off, Hamilton recorded a poverty rate of 16.7%, which was well above the average rate of Ontario (13.7%) and the average national rate (12.8%).^[https://www.hamiltoncommunityfoundation.ca/vital-signs/low-income-2018/]. According to data from Hamilton Food Share^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf], approximately 23,000 individuals accessed food banks in the city in March of 2019. Within this total is 9,125 visits by children (minors up to 18 years old), up from 8,278 the year before. Feed Ontario, the province's largest collective of hunger-relief organizations, found that on a per-capita basis, the level of need in the inner core of central Hamilton was second highest in Ontario^[https://feedontario.ca/research/report-hunger-map/].

Geographically, the "old" city of Hamilton was amalgamated with several of its surrounding municipalities in 2001, with the city now featuring a mix of urban, suburban, exurban, and rural areas. Lower-cost housing proximate to the city's industrial north end has traditionally attracted immigrants and less-affluent residents compared to the city's wealthier suburbs. However, the decentralization of population from the inner core has led to challenges in transit connectivity to amenities and services and the proportion of auto users compared to transit users remains very high [@behan2008smart; @topalovic2012light]. In addition, the city is separated geographically by the Niagara Escarpment. With sections of rocky cliff that approach 100m in height, the escarpment presents a significant challenge for promoting active travel and transport connections between "mountain" and "lower city" neighbourhoods. Taken together, the high level of food need, population locations, and transportation network characteristics combine to inform spatial accessibility to food banks and food services in the city.

@@

The locations of food banks and related food services were obtained from the Hamilton Public Library's Food Access Guide^[http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf]. The guide was updated in April of 2021 to indicate any change affected on the services due to the pandemic. This includes modified business hours, a need to make reservations before frequenting, and locations that have completely shut down in consequence. Table 1 defines each service type and the number of locations pre- and during the COVID-19 pandemic. While some food bank services have a specific a target population, such as prioritizing families with young children aged between 0 and 3 or accepting only those providing proof of low-income status through housing and utility statements, all the food bank services indicated below are designed to accommodate those in need of food at zero to low cost. With our focus on food banks and food services that offer free or low-cost meals at particular locations, we first removed services such as Meals on Wheels and other food access services such as food box, community kitchens, student nutrition programs, and shopping and transportation. With some providers offering different food services at the same location (e.g. food bank with free and community meal services), and some of these services closing after the onset of the COVID-19 pandemic, we opted to geocode based on the service type. On the other hand, two free meal services held on different days at the same location were collapsed into a single service point for the accessibility analysis. Additional details on the operations of individual facilities is not publicly available, and with the changes in operations it proved unfeasible to collect it. For this reason, the analysis to follow is of accessibility to the location of food banks and services, but not to specific services (e.g., breakfasts vs. food boxes).

@@

Food insecurity is a significant issue for many Canadian households and is associated reductions in nutrition and mental and physical health. While food bank use can provide some relief, levels of need in the City of Hamilton were on the rise before the onset of the COVID-19 pandemic with a troubling increase in the number of visits made by children. While temporary financial relief measures helped reduce food bank use in the summer of 2020, the pandemic has more than likely increased food insecurity for many households and created a surge in food bank demand in the city. On the other hand, the pandemic has also resulted in major disruptions to food bank and food service delivery, resulting in the closure or cancellation of many food services in Hamilton. On top of this, reductions in transit service have reduced travel options for many.

In response, this research has sought to better understand accessibility to food banks and food service locations as well as how the closure of some locations over the pre- and during-COVID time periods affected the potential for low-income households to reach these amenities. Using the balanced floating catchment area method to account for population demand and congestion effects at service points, we estimate multi-modal accessibility to food bank and food service locations for low-income households. In the absence of information on how food insecure households travel to food banks and related services, we adopted a normative approach that examines accessibility to food banks reachable within 15 minutes based on global standards for accessible drinking water. The weighting of travel time estimates by the modal split in different zonal geographies tailors the results to patterns of travel behaviour captured in the TTS regional travel survey. Moreover, our parcel-level analysis presents a disaggregate approach to estimating accessibility based on the locations of residential parcels and dwellings.

Results show that while accessibility levels were lower in the city's more car-oriented suburban and rural areas to begin with, the closure of 14% of the city's food bank and low-cost or free food service locations during the pandemic resulted in an overall decrease in accessibility across the city. However, these effects were not uniform over space or for different population groups. Because the balanced floating catchment area method takes into account changes in demand and congestion for service providers, the closure of food bank and food service locations can reverberate throughout the catchment areas of the city. For some suburban zones, the closure of a relatively high level-of-service location results in the remaining services being spread over a larger population. In others, high auto dependence for trips leads to decreases in accessibility that accumulate due to the loss of several food bank and service locations initially reachable within 15 minutes by car. Reductions in accessibility in the city's more urban north end, where low-income households conduct higher proportions of trips by transit and walking, emphasize the importance of geographic proximity in the potential to reach food bank and service locations for these residents. Beyond geography, the results also highlight the differential impact of food bank closing during COVID-19 on population groups with seniors and children most affected.

Still, the work is not without its limitations. We use data from the census to model low-income households as a proxy for households that are possibly experiencing food insecurity. However, the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all food banks and food service locations equally. Future work could improve this method by accounting for differential capacities at food provider locations, such as the number of meals served. Unforunately, this data was not available in the present case as some service locations had already closed. We also did not explicitly account for any changes in demand over the study period. If trends in Hamilton are similar to those in Toronto, there could be a significant increase in demand that would alter the estimated accessibilities. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown as approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations.

Nevertheless, while previous work has noted the important role of geography alongside other socioeconomic and demographic indicators in household access to healthy food, this is, to the best of our knowledge, the first study that has focused on the geographic component of food bank accessibility. Beyond accounting for the inflation of demand that occurs in traditional floating catchment area methods, our application of the balanced floating catchment area approach offers a novel analysis of food bank and food service accessibility that is sensitive to the locations of low-income households, information on their age distributions and typical tripmaking behaviour, food bank and food service locations and their operations over time, and the characteristics of the city's multi-modal transportation network. Findings from this analysis can provide quantitative evidence to help urban planners and community administrators strategically allocate food bank services considering accessibility for different population groups, especially when considering the unequal distribution of impacts from the COVID-19 pandemic.
